\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-13 paper.tex
Tweak figs
Second pass at prospective/retrospective
To address Andrew's comment, my thoughts, and (partially) Gideon's comments about application

\section*{Prospective vs retrospective ARGs}
Event ARGs, as defined in the previous section as a way of encompassing
the range of definitions provided in the literature, are
inherently \emph{retrospective}.
Event ARGs are defined in terms
of the events that occured in the history of some set of sampled
genomes.
For example, Fig.~\ref{fig-arg-in-pedigree}
is retrospective because it only includes information
about genetic inheritance relative to the sample individuals
$D_1$ and $D_2$ (genomes \noderef{a}--\noderef{d}).
That is, only the inheritance
history of the sample genomes \noderef{a}--\noderef{d} is recorded in the ARG.
% Gregor & Jerome think that "inheritance" is clearer here than "transmission"
Genetic inheritance must have occurred between (e.g.)
$D_8$ and her daughter $D_6$, but this is not recorded because
it is not relevant to the genetic ancestry of the sample genomes.

There is no particular reason, however, that ARGs must be
retrospective. They can also be \emph{prospective}.
The edges in a gARG represent a path of genetic inheritance from
ancestor to descendant through some
number of generations
% Wilder found cell-to-cell a little confusing here, we could delete
% if necessary (it's tackled later on)
(ultimately from cell to cell; see Appendix~\ref{cell-lineages-and-args}),
and are annotated
with genome coordinates over which inheritance occurs.
In a prospective gARG, the inheritance intervals $I$
on an edge joining ancestor genome $v$ and descendant $u$
defines \emph{all} regions of genome that $u$ inherited from $v$.
Thus, these inheritance intervals are
concerned only with the genetic relationship
between the nodes in question, and are not defined with respect
to any given set of samples. We are simply recording the
genetic inheritance between two genomes
(in the simplest case, between a parent and a child).

Prospective ARGs arise in individual-based forwards-time
simulations~\citep{kelleher2018efficient,haller2018tree},
where the genetic inheritance between genomes is recorded
exhaustively and periodically
``simplified'' to retain only
the ancestry relevant to the current population.
\citet{kelleher2018efficient} refer to the structure
recording all such forwards-in-time genetic relationships
as an ``embellished pedigree`` (and suggest the term ``nedigree'' as a shorthand).
Describing this instead as a ``prospective ARG'', emphasising
the use of the same data structures and the direct connection with
the usual retrospective interpretation of an ARG, may be
more useful.
The ``simplify'' algorithm, which prunes away parts of the graph that
are not ancestral to a sample, is the key enabling factor for
this approach to forwards-time simulation and is
described in detail by
\citet{kelleher2018efficient}.
It is a powerful tool for simulation and analysis as it
allows efficient subsetting of ARGs, with the ability to
remove both samples and certain levels of detail about recombination,
as we discuss in the following sections.

@@

\section{Additional plots}

\begin{figure}[h]
	\begin{center}
		\includegraphics[width=0.8\textwidth]{illustrations/simplification-with-edges.pdf}
	\end{center}
	\caption{\label{fig-simplification-with-edges}
	The four levels of ARG simplification from Fig.~\ref{fig-simplification}A--D, with edge intervals shown.
	}
\end{figure}

>>>>

\section*{Prospective vs retrospective ARGs}
The ARG concept arose from a \emph{retrospective} view of inheritance.
For example the definition of an eARG, as discussed above, is invariably
couched in terms of the events that occured in the history of some set of sampled
genomes. Similarly, the genome ARG in Fig.~\ref{fig-arg-in-pedigree}
is retrospective because it only includes information
about genetic inheritance relative to the sample individuals
$D_1$ and $D_2$ (genomes \noderef{a}--\noderef{d}).
That is, only the inheritance
history of the sample genomes \noderef{a}--\noderef{d} is recorded in the ARG.
% Gregor & Jerome think that "inheritance" is clearer here than "transmission"
Genetic inheritance must have occurred between (e.g.)
$D_8$ and her daughter $D_6$, but this is not recorded because
it is not relevant to the genetic ancestry of the sample genomes.

There is no particular reason, however, that ARGs must be
retrospective. Recent advances in individual-based forward-time simulation
~\citep{kelleher2018efficient,haller2018tree} depend upon the ability to
record gARGs in a \emph{prospective} manner.
In a prospective gARG, the inheritance intervals $I$
on an edge joining ancestor genome $v$ and descendant $u$
define \emph{all} regions of genome that $u$ inherited from $v$.
Thus, these inheritance intervals are
concerned only with the genetic relationship
between the nodes in question, and are not defined with respect
to any given set of samples. They simply record the immediate
genetic inheritance between two genomes
(in the simplest case, between a parent and a child).
\citet{kelleher2018efficient} refer to the structure
recording all such forwards-in-time genetic relationships
as an ``embellished pedigree`` (and suggest the term ``nedigree'' as a shorthand).
Describing this instead as a ``prospective ARG'', emphasising
the use of the same data structures and the direct connection with
the usual retrospective interpretation of an ARG, may be
more useful.

In a prospective ARG, the exhastive recording of all inheritance intervals from generation
to generation leads to a rapid build-up of information, much of which
may not be relevant to the genetic ancestry of the current population. 
Redundant nodes and inheritance intervals can therefore be periodically
``simplified'' to retain only
the ancestry relevant to the current population.
The ``simplify'' algorithm, which prunes away parts of the graph that
are not ancestral to a sample, is the key enabling factor for
this approach to forwards-time simulation and is
described in detail by
\citet{kelleher2018efficient}.
It is a powerful tool for simulation and analysis as it
allows efficient subsetting of ARGs, with the ability to
remove both samples and certain levels of detail about recombination,
as we discuss in the following sections.

@@

\clearpage
\section*{Supplementary figures}

% TODO add label and change numbering to S1, etc
\begin{figure}[ht]
	\begin{center}
		\includegraphics[width=0.8\textwidth]{illustrations/simplification-with-edges.pdf}
	\end{center}
	\caption{\label{fig-simplification-with-edges}
	The four levels of ARG simplification from Fig.~\ref{fig-simplification}A--D, with edge intervals shown.
	}
\end{figure}
